---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Atlas

Automated model building for R, driven by an agentic LLM
([ellmer](https://ellmer.tidyverse.org)). Give Atlas a dataset and a target;
it explores the data, proposes a modeling plan for your approval, then writes
and runs R code to fit and evaluate candidate models — pausing for your input
when a decision needs you. You get back fitted models, a leaderboard, the
full code trail, and a report of how each model was built.

**Your data stays on your machine.** The LLM sees a compact profile of the
data and whatever output the agent prints — never the rows themselves. All
model fitting happens locally in your R session.

## Installation

```{r}
# install.packages("pak")
pak::local_install("path/to/Atlas")
```

Set an API key for your provider (default: `ANTHROPIC_API_KEY`). The
persistent way: `usethis::edit_r_environ()`, add
`ANTHROPIC_API_KEY=sk-ant-...`, save, restart R. Any tool-capable ellmer
provider works via the `chat` argument.

## Quickstart

```{r}
library(Atlas)

res <- atlas(mtcars, outcome = "mpg",
             n_models = 3,        # maximum; stops earlier once converged
             patience = 3,        # stop after 3 attempts without improvement
             min_improve = 0.05,  # <5% better doesn't count as improvement
             goal = "prioritise interpretability")

res              # leaderboard + report of how each model was built
res$models       # named list of fitted models, ready for predict()
res$code         # every line of R the agent ran
```

After the winning algorithm is found, Atlas keeps refining its feature
selection and engineering (one change per attempt, same validation scheme)
until the stopping rules say it has converged — the optimum lands in the
results as `<winner>_refined`. The session stays live for anything further:

```{r}
res$session$tell("why did the refined model beat the original?")
res$session$tell("try a challenger that uses at most 3 predictors")
```

## Domain knowledge as enforced constraints

Constraints are hard requirements, and machine-checked ones are *verified*
against every final model — violations are sent back to the agent to fix:

```{r}
res <- atlas(mtcars, "mpg", constraints = list(
  uses_wt = con_uses("wt"),                   # must use wt
  mono_hp = con_monotone("hp", "decreasing"), # monotone in hp
  no_leak = "never use qsec; it is measured after the fact"
))
res$constraints   # compliance table: model x constraint
```

Or write the brief in plain language and let the LLM structure it:

```{r}
cons <- extract_constraints(
  "weight must be in the model; predictions can never go up as
   horsepower rises", data = mtcars)
```

## Persistent sessions

Every run checkpoints to its own directory: the data, each code chunk, the
conversation, and the final artifacts (reports, validation plots, model
bundles). By default runs land in `.atlas/<timestamp>` under the working
directory — set `options(atlas.dir = "~/atlas-runs")` in your `.Rprofile`
to send them anywhere you like. Sessions survive crashes and restarts:

```{r}
s <- atlas_resume(".atlas/20260703-141500")
s$tell("continue where you left off")
```

## How big can the data go?

Rows are cheap, columns are the limit that matters:

* **Rows** — bounded by your machine, not the LLM: the data is never sent to
  the model, and fitting happens locally. Hundreds of thousands to millions
  of rows work; model fitting and constraint checks just take longer.
* **Columns** — the practical LLM-side limit. The agent reasons about
  variables by name; up to a few dozen columns it does that well, into the
  low hundreds it degrades. For wide data, pre-select features first or say
  how to in `goal`.
* **Cost & context** — a typical build is a few dozen LLM calls; cost scales
  with the number of agent steps (and `max_fix_rounds` repairs), not with
  data size. Start a fresh session per modeling task rather than piling many
  follow-ups into one conversation.
* **Disk** — the run directory stores a copy of the data (`data.rds`) so
  sessions can be resumed cold; budget for that with very large datasets.

## Learn more

* `vignette("atlas")` — getting started, API keys, what you get back
* `vignette("constraints")` — domain knowledge, custom checks, NL extraction
* `vignette("sessions")` — persistence, resuming, follow-ups
