---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Atlas

Atlas builds models for you, and shows its work. Give it a data frame and a
target; an LLM agent (via [ellmer](https://ellmer.tidyverse.org)) explores
the data, proposes a modeling plan for your approval, then writes and runs R
code to fit, compare, and refine candidate models. You get back fitted
models ready for `predict()`, a leaderboard, a written report — and, because
the agent works by writing R, the complete script of everything it did.

Three ideas separate Atlas from "ask a chatbot for model code":

* **Domain knowledge is enforced, not suggested.** Constraints like "`wt`
  must be used" or "predictions must never rise with horsepower" are
  machine-verified against every final model, and violations are sent back
  to the agent to fix.

* **Sessions are persistent.** Every code chunk and conversation turn is
  checkpointed to disk as it happens. A crash, a restart, or a deliberate
  interruption loses nothing — resume and keep going.

* **Your data stays on your machine.** The LLM sees column names, types,
  and printed summaries, never the rows. All fitting happens locally in
  your R session.

## Installation

Atlas is not yet on CRAN. Install from a local checkout or GitHub:

```{r}
# install.packages("pak")
pak::local_install("path/to/Atlas")
```

You also need an API key for an LLM provider. Atlas defaults to Anthropic:
run `usethis::edit_r_environ()`, add a line `ANTHROPIC_API_KEY=sk-ant-...`,
save, and restart R. Any tool-capable ellmer provider works via the `chat`
argument.

## Usage

```{r}
library(Atlas)

res <- atlas(mtcars, outcome = "mpg", goal = "prioritise interpretability")
```

The agent narrates as it works. It profiles the data (flagging suspected
target leakage), proposes a plan — model families chosen to match the
outcome's distribution, a validation scheme, any monotonicity it thinks the
domain implies — and **waits for your approval in the console**. You can
answer "yes", or redirect it: "only linear models, and don't use qsec".
Then it builds up to `n_models` candidates, compares them on held-out data,
refines the winner's features until improvement stalls, and verifies any
constraints.

```{r}
res               # leaderboard, constraint compliance, report
res$models        # fitted models: predict(res$models$glm1, newdata)
res$code          # every line of R the agent ran, in order
res$session$tell("why did the refined model win?")   # keep talking
```

Convergence is under your control: `n_models` is a maximum, and the
stopping rules — `patience` attempts without improvement, gains under
`min_improve` don't count — govern both candidate building and refinement.

## Constraints

Pass hard requirements as a list. Strings are enforced through instructions;
constraint objects carry a `check` function that Atlas runs against every
final model, feeding failures back to the agent for repair:

```{r}
res <- atlas(mtcars, "mpg", constraints = list(
  uses_wt = con_uses("wt"),                    # must actually use wt
  mono_hp = con_monotone("hp", "decreasing"),  # effect must be monotone
  no_leak = "qsec is measured after the fact; never use it"
))
res$constraints    # one row per model x constraint: passed, detail
```

Columns that won't exist at prediction time don't belong in the data at
all — `exclude` removes them before the agent ever sees it, and
`atlas_leakage_screen()` automatically flags predictors that alone explain
almost all of the outcome. See `vignette("constraints")`.

## Sessions

`atlas()` is a one-call wrapper around an `AtlasSession`, which lives in the
run directory as much as in memory: data, code log, conversation, and
artifacts (report, leaderboard, models, validation plots) are all on disk,
so sessions survive anything:

```{r}
s <- atlas_resume(".atlas/20260703-141500")
s$tell("swap the elastic net for a GAM and re-compare")
```

Runs land in `.atlas/<timestamp>` by default; set
`options(atlas.dir = "~/atlas-runs")` to choose. See `vignette("sessions")`.

## Limitations

* **Columns are the scaling limit, not rows.** Fitting is local, so a
  million rows just take the time they take — but the agent reasons about
  variables by name, and past a few dozen columns that reasoning degrades.
  Pre-select features for wide data.
* **Cost scales with agent steps**, not data size: more candidates, more
  repair rounds, more follow-ups mean more LLM calls.
* **An agent is not a statistician.** Atlas verifies what you tell it to
  verify; judgment about what the model is *for* stays with you. Read the
  report, check `res$code`, and look at the validation output.

## Learn more

* `vignette("atlas")` — a full walkthrough: setup, the build lifecycle,
  what you get back
* `vignette("constraints")` — encoding domain knowledge that can't be
  ignored
* `vignette("sessions")` — persistence, resuming, and steering mid-build
