---
title: "Getting started with Atlas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Atlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Atlas automates the workmanlike part of modeling — explore, propose, fit,
compare, refine, document — while keeping you in charge of every judgment
call. An LLM agent does the work by writing and running R code in your
session, narrating as it goes and pausing for your approval at the moments
that matter.

This vignette walks through one build end to end. Code chunks are not
evaluated here because they need an API key.

Two facts worth internalising before the first run:

* **Your data never leaves your machine.** The agent receives a compact
  profile — dimensions, column names and types — plus whatever output its
  own code prints. Model fitting is ordinary local R.
* **Everything is recorded.** Every code chunk the agent executes and every
  conversational turn is checkpointed to a run directory as it happens.
  The result of a build is not just models; it is a reproducible script.

## Setup

Atlas talks to the LLM through [ellmer](https://ellmer.tidyverse.org) and
defaults to Anthropic's API, which reads a key from the `ANTHROPIC_API_KEY`
environment variable. The tidy, persistent way to set it:

```{r}
usethis::edit_r_environ()
```

Add one line (no quotes), save, restart R:

```
ANTHROPIC_API_KEY=sk-ant-...
```

Every tool-capable ellmer provider is a drop-in replacement — pass one via
`chat`:

```{r}
res <- atlas(mtcars, "mpg", chat = ellmer::chat_openai(model = "gpt-5"))
```

## One build, start to finish

```{r}
library(atlas)

res <- atlas(mtcars, outcome = "mpg", n_models = 3,
             goal = "prioritise interpretability")
```

Here is the lifecycle that call runs, and where you come in:

1.  **Explore.** The agent profiles the data: dimensions, types,
    missingness, and the outcome's distribution. Predictors that alone
    explain nearly all of the outcome are flagged as suspected target
    leakage (see `atlas_leakage_screen()`) and must be confirmed with you
    before the agent may use them.

2.  **Plan — and stop for you.** The agent proposes candidate model
    families *chosen to match the outcome's distribution* (binary outcome
    → binomial; counts → Poisson-family; skewed positive → Gamma/Tweedie
    or a log transform), a validation scheme with a fixed seed, and any
    monotone effects it believes the domain implies. It then asks for your
    approval in the console. This is a conversation, not a checkbox: answer
    `yes`, or redirect — "drop the tree model, and don't use qsec" — and
    the plan is revised before anything is fitted.

3.  **Build.** Up to `n_models` candidates are fitted and compared on
    held-out data, one narrated line per candidate.

4.  **Refine.** The winner's feature selection and engineering are iterated
    — one change per attempt, same validation scheme — until the stopping
    rules call convergence. The refined model joins the results as
    `<winner>_refined`, next to the original.

5.  **Verify.** Machine-checked constraints (if you passed any) are tested
    against every final model; violations go back to the agent for repair.
    See `vignette("constraints")`.

6.  **Document.** With the `modelblueprint` package installed, the winning
    model gets a full validation workup — gain chart, calibration, grouped
    residuals, one-way plots, PDPs — written to the run directory as
    interactive HTML, alongside a portable model bundle.

## Stopping rules

`n_models` is a maximum, not a quota. Two numeric parameters decide when
the agent should stop iterating, and they govern candidate building and
refinement alike:

```{r}
res <- atlas(mtcars, "mpg",
  n_models           = 5,     # at most five candidates
  stopping_rounds    = 3,     # stop after 3 attempts without improvement
  stopping_tolerance = 0.05)  # gains under 5% (relative) don't count
```

The report always states *why* iteration stopped — limit reached or
converged. These rules are applied by the agent itself; for hard
guarantees on unattended runs there are also mechanically enforced
budgets, `max_steps` (code executions) and `max_runtime` (seconds) —
once exhausted, the execution tool refuses to run further code and the
agent must finalise from what it has.

## What you get back

```{r}
res                    # print: leaderboard, constraint status, report
res$models             # named list of fitted models
predict(res$models[[1]], head(mtcars))
res$leaderboard        # validation metric per model, best first
res$report             # markdown: how each model was built, and why
res$constraints        # compliance table (if constraints were set)
cat(res$code, sep = "\n\n")   # the full script the agent ran
res$dir                # the run directory holding all of the above
```

One element deserves special mention: `res$session` is the live session
object, with the full conversation still in context. Anything you would ask
a colleague who just built these models, you can ask it:

```{r}
res$session$tell("why did the refined model beat the original?")
res$session$tell("build one more candidate that uses at most 3 predictors")
```

Sessions also survive R itself — see `vignette("sessions")` for resuming,
crash recovery, and interrupting a build in progress.

## How big can the data be?

Rows are cheap; columns are the limit that matters.

* **Rows: bounded by your machine, not the LLM.** The data is never
  uploaded, and printed tool output is capped (~8,000 characters per
  chunk), so a million rows cost you fitting time, not tokens.
* **Columns: double digits are ideal, low hundreds the practical ceiling.**
  The agent reasons about variables by name; its data profile lists up to
  50 columns. For wide data — one-hot blocks, text features, omics —
  pre-select first, or tell it how in `goal` ("variance filter, then work
  with the top 30").
* **Cost scales with agent steps.** More candidates, more repair rounds,
  and more follow-ups mean more LLM calls; more rows don't. A typical
  build is a few dozen calls.
* **Disk:** the run directory keeps a copy of the data so sessions resume
  cold. Point `dir` (or `options(atlas.dir = )`) somewhere with room for
  large datasets.
