---
title: "Building models with Atlas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building models with Atlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Atlas hands your data to an LLM agent that explores it, proposes a modeling
plan for your approval, then fits and evaluates candidate models by writing
and running R code. Chunks below are not evaluated at build time because they
need an API key (`ANTHROPIC_API_KEY` by default, or any ellmer provider).

## Setting your API key

Atlas talks to the LLM through ellmer, which reads the key from an
environment variable — `ANTHROPIC_API_KEY` for the default provider. Without
it you'll see:

```
Error in `anthropic_key()`:
! Can't find env var `ANTHROPIC_API_KEY`.
```

The recommended, persistent way is to put it in your `.Renviron`:

```{r}
usethis::edit_r_environ()
```

add this line (no quotes), save, then restart R:

```
ANTHROPIC_API_KEY=sk-ant-...
```

For a one-off session you can set it inline instead (don't save this in a
script you share):

```{r}
Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-...")
```

Other providers use their own variables, e.g. `OPENAI_API_KEY` for
`ellmer::chat_openai()` — see the [ellmer provider
docs](https://ellmer.tidyverse.org/reference/index.html#chatbots) for the
full list.

## One call

```{r}
library(Atlas)

res <- atlas(mtcars, outcome = "mpg", n_models = 1,
             goal = "prioritise interpretability")
```

The agent streams what it is doing, and pauses in the console when it wants
your approval of the plan or input on a decision. When it finishes:

```{r}
res                  # leaderboard + markdown report of how each model was built
res$models           # named list of fitted models
predict(res$models[[1]], head(mtcars))
cat(res$code, sep = "\n\n")   # every line of R the agent ran
```

## Domain knowledge: constraints

`goal` carries soft preferences; `constraints` carry hard requirements. Each
constraint is a plain-language rule the agent must follow, optionally with a
`check` function — and checked constraints are *verified*: after the build,
every final model is tested, and violations are sent back to the agent to fix
(up to `max_fix_rounds` times).

```{r}
res <- atlas(mtcars, "mpg", n_models = 2, constraints = list(
  uses_wt = con_uses("wt"),                  # wt must contribute to predictions
  mono_hp = con_monotone("hp", "decreasing"), # predictions decreasing in hp
  no_leak = "qsec is measured after the fact; never use it as a predictor"
))

res$constraints   # compliance table: model x constraint, passed, detail
```

The built-in helpers are model-agnostic (they only need
`predict(model, newdata)`): `con_uses()` scrambles the column and checks that
predictions move; `con_monotone()` sweeps the variable over its range with
other predictors held fixed and checks the direction of every step.

Custom checks are just functions of `(model, data)` returning `TRUE` or a
description of the violation:

```{r}
non_negative <- constraint(
  "predicted mpg must never be negative",
  check = function(model, data) {
    p <- predict(model, newdata = data)
    if (all(p >= 0)) TRUE else "some predictions are negative"
  }
)
```

Plain strings (like `no_leak` above) are enforced through instructions and
shown as `passed = NA` in the compliance table — use a `check` whenever the
rule can be tested in code.

## Sessions: follow-ups and persistence

`atlas()` is a wrapper around an `AtlasSession`, which stays alive for
follow-up work in the same context:

```{r}
res$session$tell("why did the random forest win?")
res$session$tell("add a 4th candidate using only 3 predictors")
```

Everything is checkpointed to a run directory (`.atlas/<timestamp>` by
default): the data, every code chunk, the full conversation, and the final
artifacts (`report.md`, `leaderboard.csv`, `models.rds`). If R crashes or you
come back tomorrow, resume:

```{r}
s <- atlas_resume(".atlas/20260703-141500")
s$tell("continue where you left off")
s$results()
```

Resume replays the recorded code log against the saved data, so the working
environment — fitted models included — is reconstructed exactly.

## Choosing a provider

Any tool-capable ellmer chat works:

```{r}
res <- atlas(mtcars, "mpg", chat = ellmer::chat_openai(model = "gpt-5"))
```
