---
title: "Getting started with Atlas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Atlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Atlas hands your data to an LLM agent that explores it, proposes a modeling
plan for your approval, then writes and runs R code to fit and evaluate
candidate models — pausing in the console when a decision needs you. Chunks
in this vignette are not evaluated at build time because they need an API
key.

Your data is never uploaded: the LLM sees a compact profile (column names,
types, dimensions) and whatever output the agent prints. All code runs
locally in your R session.

## Setting your API key

Atlas talks to the LLM through ellmer, which reads the key from an
environment variable — `ANTHROPIC_API_KEY` for the default provider. Without
it you'll see:

```
Error in `anthropic_key()`:
! Can't find env var `ANTHROPIC_API_KEY`.
```

The recommended, persistent way is to put it in your `.Renviron`:

```{r}
usethis::edit_r_environ()
```

add this line (no quotes), save, then restart R:

```
ANTHROPIC_API_KEY=sk-ant-...
```

For a one-off session you can set it inline instead (don't save this in a
script you share):

```{r}
Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-...")
```

Other providers use their own variables, e.g. `OPENAI_API_KEY` for
`ellmer::chat_openai()` — see the [ellmer provider
docs](https://ellmer.tidyverse.org/reference/index.html#chatbots) for the
full list. Any tool-capable provider works:

```{r}
res <- atlas(mtcars, "mpg", chat = ellmer::chat_openai(model = "gpt-5"))
```

## No coding required

If you (or the person you're handing this to) would rather not write R at
all, launch the app:

```{r}
Atlas::atlas_app()
```

It opens in the browser: upload a CSV, pick the column to predict, optionally
type goals and plain-English rules (turned into enforced constraints via
`extract_constraints()`), and click **Build models**. Progress streams live,
questions from the agent appear as pop-up dialogs, and the results tabs show
the leaderboard, the rules check, the report, and the complete R code of the
build. Everything is also saved to the run directory as usual, so an R user
can pick the session up later with `atlas_resume()`.

The app needs the `shiny` and `callr` packages
(`install.packages(c("shiny", "callr"))`) and the same API key as the
console workflow.

## One call

```{r}
library(Atlas)

res <- atlas(mtcars, outcome = "mpg", n_models = 1,
             goal = "prioritise interpretability")
```

The agent streams what it is doing — narration, each code chunk under an
`── R ──` rule with its output as `#>` lines — and pauses in the console when
it wants your approval of the plan or input on a decision.

## What you get back

```{r}
res                  # leaderboard + markdown report of how each model was built
res$models           # named list of fitted models
predict(res$models[[1]], head(mtcars))
res$leaderboard      # validation metric per model, best first
res$report           # markdown: how each model was built, recommendation
cat(res$code, sep = "\n\n")   # every line of R the agent ran
res$constraints      # compliance table, if you passed constraints
res$session          # the live session, for follow-ups
```

The session is the persistent part — follow up in the same context:

```{r}
res$session$tell("why did this model win?")
res$session$tell("add another candidate using only 3 predictors")
```

See `vignette("sessions")` for persistence, crash recovery, and resuming, and
`vignette("constraints")` for passing domain knowledge as enforced
constraints.

## Refinement: from best algorithm to optimum model

Finding the best algorithm is only half the job. By default (`refine =
TRUE`), once the candidates are compared and constraints pass, Atlas takes
the winner and iterates on its feature selection and engineering — one
change per attempt (add/drop predictors, transformations, interactions,
binning), refit, evaluate on the same validation scheme — until the stopping
rules trigger: `patience` consecutive attempts without a relative gain of at
least `min_improve`. The refined model is added to the results as
`<winner>_refined` next to the original, constraints are re-verified against
it, and the report says what improved and why iteration stopped.

## Reviewable validation output

If the [modelblueprint](https://mattyoreilly.github.io/modelblueprint/)
package is installed, every build ends with a validation step (`validate =
TRUE`, the default): the agent wraps the winning model in a `modelblueprint`
and runs `model_validation()`, writing interactive Plotly HTML files — gain
chart, predicted-vs-observed calibration, grouped residuals, one-way plots
per feature, and PDPs — into the run directory for review. In the app these
appear as clickable links in a "Validation plots" card on the Results tab.

The plan the agent proposes is also required to be driven by the outcome's
distribution — binary outcomes get binomial models, counts get
Poisson-family, skewed positive targets get Gamma/Tweedie or a log transform
— with the reasoning stated in the plan you approve.

## How big can the data be?

Rows are cheap; columns are the limit that matters.

* **Rows: bounded by your machine, not the LLM.** The data itself is never
  sent to the model — only a profile and printed summaries — and all fitting
  runs locally. A million rows is fine from Atlas's side; model fitting,
  cross-validation, and constraint checks simply take as long as they would
  if you ran them yourself. Tool output shown to the agent is capped (~8000
  characters per chunk), so printing a huge object costs tokens but does not
  explode.
* **Columns: keep it double digits, low hundreds at most.** The agent
  reasons about variables by name. The data profile it receives lists up to
  50 columns; with a few dozen columns it can think about each one, and into
  the hundreds it has to work with aggregates. For wide data (text features,
  one-hot blocks, omics), pre-select features first — or tell it how to in
  `goal` ("run a variance filter, then work with the top 30").
* **Conversation length.** A typical build is 15–40 LLM steps, comfortably
  inside the model's context window. Many follow-ups and repair rounds
  accumulate, though — for a new modeling question, start a fresh
  `atlas()` call instead of extending an old session forever.
* **Cost scales with agent steps, not data size.** More candidate models,
  more constraint repairs, and more follow-ups mean more LLM calls; more
  rows don't.
* **Disk.** The run directory keeps a copy of the data (`data.rds`) so
  sessions resume cold; with very large datasets, point `dir =` somewhere
  with room.
