---
title: "Domain knowledge and constraints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Domain knowledge and constraints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`goal` carries soft preferences ("prioritise interpretability"); `constraints`
carry hard requirements that must hold for every final model. Atlas enforces
them three ways:

1. **Up front** — constraints go into the agent's task as numbered hard
   requirements, and the agent is told to design for them from the start
   (e.g. pick monotone model forms rather than fixing violations later).
2. **During the build** — the agent has a `check_constraints` tool to verify
   itself before declaring the build done.
3. **After the build** — Atlas independently runs every machine check against
   every model in the results. Violations are sent back to the agent to
   refit, up to `max_fix_rounds` times (default 2).

## Three ways to write a constraint

```{r}
library(Atlas)

res <- atlas(mtcars, "mpg", n_models = 2, constraints = list(
  # 1. built-in helpers: machine-checked, model-agnostic
  uses_wt = con_uses("wt"),
  mono_hp = con_monotone("hp", "decreasing"),

  # 2. custom check: any function(model, data) -> TRUE or a violation message
  non_neg = constraint(
    "predicted mpg must never be negative",
    check = function(model, data) {
      p <- predict(model, newdata = data)
      if (all(p >= 0)) TRUE else "some predictions are negative"
    }
  ),

  # 3. plain string: enforced via instructions only (passed = NA in the table)
  no_leak = "qsec is measured after the fact; never use it as a predictor"
))

res$constraints
#>   model constraint passed detail
#>   ...     one row per model x machine-checked constraint ...
```

Anything that can be tested in code should get a `check`; strings are for
rules only the agent can honour (like "don't use this column"). A check may
also `stop()` — errors count as violations, with the message as the detail.

## The built-in helpers

Both are model-agnostic: they only require
`predict(model, newdata = <data.frame>)` to work, which Atlas tells the agent
every final model must support.

**`con_uses(var)`** — the model must actually use a variable. The check
scrambles (reverses) the column and compares predictions: if they don't move,
the model isn't using it. No formula inspection, so it works for any model
class and can't be fooled by wrappers.

**`con_monotone(var, direction)`** — predictions must be monotone in a
variable. ICE-style check: for a sample of observed rows, the variable is
swept across its observed range with all other predictors held fixed, and
every step of every sweep must move in the stated direction (within `tol`).

Monotonicity is *non-strict*, so a model that ignores the variable passes
trivially (flat is monotone). That makes it a conditional constraint — "if
the model responds to `hp`, the effect must be decreasing" — without forcing
the variable in:

```{r}
# monotone if present:
constraints = list(mono_hp = con_monotone("hp", "decreasing"))

# must be used AND monotone:
constraints = list(uses_hp = con_uses("hp"),
                   mono_hp = con_monotone("hp", "decreasing"))
```

## More custom check patterns

```{r}
# accuracy floor on a reference set
min_r2 <- constraint(
  "in-sample R-squared must be at least 0.7",
  check = function(model, data) {
    r2 <- cor(predict(model, newdata = data), data$mpg)^2
    if (r2 >= 0.7) TRUE else sprintf("R-squared is %.2f", r2)
  }
)

# bounded predictions
in_range <- constraint(
  "predictions must stay within the observed outcome range",
  check = function(model, data) {
    p <- predict(model, newdata = data)
    if (all(p >= min(data$mpg) & p <= max(data$mpg))) TRUE
    else "predictions leave the observed range"
  }
)

# parity across a group
fair_gap <- constraint(
  "mean absolute error must be within 1 unit across am groups",
  check = function(model, data) {
    err <- abs(data$mpg - predict(model, newdata = data))
    gap <- abs(diff(tapply(err, data$am, mean)))
    if (gap <= 1) TRUE else sprintf("MAE gap across am is %.2f", gap)
  }
)
```

## Excluding features and catching leakage

Constraints govern *how* variables are used; `exclude` removes them entirely
— for columns that won't exist at prediction time in deployment, or that
leak the outcome. Excluded columns are dropped from the data before the
agent ever sees it, which is a stronger guarantee than any instruction:

```{r}
res <- atlas(claims, "severity",
             exclude = c("settlement_amount",   # known after the outcome
                         "claim_ref"))          # ID, useless in deployment
```

The surviving predictors are screened automatically with
`atlas_leakage_screen()` — a fast univariate check: any column that alone
explains almost all of the outcome (R-squared at/above 0.95) is flagged in
the agent's brief, and the agent must confirm with you that it is genuinely
available at prediction time before using it. You can also run the screen
yourself before any modeling:

```{r}
atlas_leakage_screen(claims, "severity")
#>            variable    r2 flagged
#>   settlement_amount 0.998    TRUE
#>            exposure 0.310   FALSE
```

## Suggested by the agent

You may forget a monotonicity constraint that's obviously right for the
domain (price should not fall as floor area grows). Atlas is instructed to
think about this during data exploration: when it believes a predictor should
have a monotone effect, it includes the suggestion — variable, direction, and
a one-line why — in the plan it asks you to approve. Each suggestion you
approve is added via its `add_monotone_constraint` tool and becomes a hard,
machine-checked constraint, verified and repaired exactly like one you passed
yourself (and persisted, so it survives resumes).

## Constraints from natural language

`extract_constraints()` parses a plain-language brief into constraint
objects, mapping to machine checks wherever it recognises a checkable rule:

```{r}
cons <- extract_constraints(
  "weight must be in the model; predictions can never go up as horsepower
   rises; qsec is measured after the fact, never use it",
  data = mtcars
)
cons
#> $uses_wt
#> [machine-checked] The variable `wt` must be used as a predictor.
#> $mono_hp
#> [machine-checked] Predictions must be monotonically decreasing in `hp`.
#> $no_qsec
#> [prompt-only]     Never use qsec as a predictor.

res <- atlas(mtcars, "mpg", constraints = cons)
```

The parse is a judgment call, so it returns the list for you to inspect and
edit before the build. Passing `data` lets it validate column names; anything
unmappable (unknown column, monotone rule without a direction) falls back to
prompt-only enforcement instead of producing a broken check.

## Reading the compliance table

`res$constraints` has one row per model x machine-checked constraint, plus
one row per prompt-only constraint:

* `passed = TRUE/FALSE` — the check ran; `detail` explains failures.
* `passed = NA` — prompt-only: enforced via instructions, not verifiable.

`print(res)` shows an `UNMET CONSTRAINTS` block if anything failed after all
repair rounds; a session's current state can be re-checked any time with
`res$session$check()`.
