---
title: "Persistent sessions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Persistent sessions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A model build is a long-running, interruptible collaboration, so Atlas
treats session state as something that lives on disk first and in memory
second. The design has one load-bearing idea: **the code log is the source
of truth**. Arbitrary R environments can't be reliably serialised, but a
recorded script re-run against the same data reconstructs one exactly.
Atlas therefore checkpoints every code chunk the moment it executes, and
"resuming" means replaying that script - models included - then restoring
the conversation.

## The session object

`atlas()` is a one-call wrapper around an `atlas_session` (R6), which holds
the ellmer conversation, the R environment the agent works in, and the run
directory:

```{r}
library(atlas)

s <- atlas_session$new(mtcars, outcome = "mpg", n_models = 3,
                      goal = "prioritise interpretability")
res <- s$build()                 # the full lifecycle, approval included
s$tell("swap model 2 for a GAM") # follow-up, same context
s$check()                        # re-verify constraints right now
s$results()                      # assemble a fresh results object
```

Every results object carries its live session as `res$session`, so a
follow-up is never more than one call away.

## What's on disk

Each session owns a run directory - by default a timestamped folder under
`getOption("atlas.dir", ".atlas")`; set that option once, or pass `dir`
per run.

| file              | contents                                            |
|-------------------|-----------------------------------------------------|
| `data.rds`        | the training data, so a resume works cold           |
| `meta.rds`        | outcome, stopping rules, goal, constraints          |
| `code.R`          | every chunk the agent ran - a readable script       |
| `code.rds`        | the same chunks, exactly as recorded, for replay    |
| `turns.rds`       | the full conversation                               |
| `report.md`       | the agent's report                                  |
| `leaderboard.csv` | validation metrics per model                        |
| `models.rds`      | the fitted models                                   |

Checkpoints are written after every tool call and every reply, so at worst
a crash loses the single in-flight step. `code.R` doubles as the
reproducibility artifact: plain R you can read, audit, or run without
Atlas - or an API key.

## Resuming

```{r}
s <- atlas_resume(".atlas/20260703-141500")
s$tell("continue where you left off")
```

`atlas_resume()` reloads the data, replays the recorded code log top to
bottom (agent code is seeded, so the replay is exact), and restores the
conversation turns. The agent comes back knowing everything it did, with
every fitted object rebuilt. The one cost to budget for: replay refits the
models, so resuming a session whose models take minutes to fit takes those
minutes again.

## Interrupting a build

You don't have to wait for the agent to ask you something. Interrupt with
`Ctrl+C` / `Esc`, then resume with the new information - the work up to the
interruption was already checkpointed:

```{r}
res <- atlas(big_data, "claim_cost")   # Ctrl+C mid-build...

s <- atlas_resume(".atlas/20260703-141500")
s$tell("stop trying tree models; the deployment target only supports GLMs")
```

Programmatic front-ends get a cleaner hook: the `interject` constructor
argument, a `function()` polled after every tool call. Return a message to
have it delivered with the agent's next tool result, marked as the
highest-priority instruction; return `NULL` when there is nothing to say.

## Staying inside the token budget

The conversation is the only part of a session that grows - and LLM APIs
re-read the whole context on every request, so an unmanaged conversation
costs more with every step and eventually overflows the model's window.
atlas manages it automatically, exploiting the same design idea as resume:
the conversation is not the real memory.

Past `compact_at` input tokens (default 100,000, cached tokens included),
the next message triggers compaction: the transcript is archived to the run
directory (`turns-archive-01.rds`, `-02`, …- nothing is deleted), the
window is cleared, and the message is prefixed with a re-orientation
briefing built from ground truth - the models in `atlas_models`, the
current leaderboard, the number of code chunks run. Because the briefing is
assembled deterministically from session state, it costs no extra LLM call
and cannot hallucinate, unlike a model-written summary.

You can also compact by hand before a long follow-up:

```{r}
res$session$compact()
res$session$tell("now write an extended technical appendix for the report")
```

Cost stays visible: `res$cost` (and the last line of `print(res)`) is the
session's cumulative dollar cost, straight from ellmer's token accounting.

## Questions from the agent

The agent asks for plan approval - and anything else it genuinely needs -
through its `ask_user` tool. In an interactive session the question is
printed and the console blocks on your answer; treat it as a conversation,
not a yes/no gate, since whatever you type is folded into the plan. In
non-interactive contexts (`Rscript`, CI) the tool tells the agent to use
its best judgment and record the decision in the report, so scripted runs
never hang. Front-ends can supply their own handler via `on_ask`.

## Verbosity

`verbose = TRUE` (default) streams the narration, each code chunk, its
output, and constraint-check tables to the console. `verbose = FALSE` runs
silently - everything still checkpoints, so `report.md` and `code.R` tell
the story afterwards.
