---
title: "Persistent sessions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Persistent sessions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`atlas()` is a one-call wrapper around an `AtlasSession` — an R6 object that
holds the conversation, the R environment the agent works in, and a run
directory on disk. The session outlives the call: use it for follow-ups, and
resume it after a crash or restart.

## The session object

```{r}
library(Atlas)

s <- AtlasSession$new(mtcars, outcome = "mpg", n_models = 3,
                      goal = "prioritise interpretability")
res <- s$build()                 # the full agentic loop
s$tell("swap model 2 for a GAM") # follow-up in the same context
s$check()                        # re-verify constraints right now
s$results()                      # assemble a fresh results object
```

`res$session` from `atlas()` is the same thing — every result carries its
live session, so `res$session$tell("why did model 1 win?")` continues the
conversation with full memory of the build.

Iterative follow-ups inherit the session's stopping rules (`patience`
attempts without improvement; improvements under the `min_improve` fraction
don't count), so "keep refining the features until it stops improving"
terminates on its own:

```{r}
s <- AtlasSession$new(mtcars, "mpg", n_models = 3,   # n_models is a maximum
                      patience = 3, min_improve = 0.05)  # 0.05 = 5%
res <- s$build()
s$tell("take the best model and refine its feature engineering until
        performance stops improving")
```

In the app, the same thing is the "Ask for more" box that appears in the
sidebar after a build finishes; it resumes the session in a fresh background
process and streams into the Progress tab.

## What's on disk

Each session owns a run directory (default `.atlas/<timestamp>`, or pass
`dir =`):

| file              | contents                                             |
|-------------------|------------------------------------------------------|
| `data.rds`        | the training data, so a resume works cold            |
| `meta.rds`        | outcome, n_models, goal, constraints                 |
| `code.R`          | every code chunk the agent ran, human-readable       |
| `code.rds`        | the same chunks, exactly as recorded (for replay)    |
| `turns.rds`       | the full conversation                                |
| `report.md`       | the agent's final report                             |
| `leaderboard.csv` | validation metrics per model                         |
| `models.rds`      | the fitted models                                    |

Checkpoints are written after every tool call and every reply, so at most the
in-flight step is lost to a crash. `code.R` doubles as the reproducibility
artifact: it is plain R you can read, audit, or run without Atlas.

## Resuming

```{r}
s <- atlas_resume(".atlas/20260703-141500")
s$tell("continue where you left off")
```

Resume does not try to deserialize the working environment. It reloads the
data, **replays the recorded code log** top to bottom (agent code is seeded,
so this is exact), and restores the conversation turns — the agent comes back
knowing everything it did, with all fitted objects rebuilt. The one thing to
budget for: replay refits the models, so resuming a session whose models take
minutes to fit takes those minutes again.

## Interrupting a running build

You don't have to wait for the agent to ask. In the app, a **"Tell Atlas
something now"** box is available in the sidebar whenever a build is
running: your message is delivered with the agent's next step, marked as its
highest-priority instruction ("I forgot — CREDIT_GRADE is only available for
existing customers"), and the agent acknowledges it before continuing. A
message sent in the instant a build finishes is returned to you with a
prompt to use "Ask for more" instead.

In the console, interrupt with `Ctrl+C` / `Esc`, then resume with the new
information — because every code chunk is checkpointed as it runs, the
agent's work up to the interruption is preserved and replayed:

```{r}
res <- atlas(big_data, "claim_cost")   # Ctrl+C mid-build...
s <- atlas_resume(".atlas/20260703-141500")
s$tell("stop trying tree models; the deployment target only supports GLMs")
```

For programmatic front-ends, the underlying hook is the `interject`
constructor argument: a `function()` polled after every tool call, returning
a message or `NULL`. The app implements it as a file in the run directory.

## Human-in-the-loop

The agent has an `ask_user` tool. In an interactive R session it prints the
question under an `── atlas needs your input ──` rule and blocks on the
console — that's how plan approval works. In non-interactive contexts
(`Rscript`, CI) the tool tells the agent to use its best judgment and record
the decision in the final report, so scripted runs never hang.

## Verbosity

`verbose = TRUE` (default) streams the agent's narration, each code chunk
under an `── R ──` rule, its output as `#>` lines, and constraint-check
tables. `verbose = FALSE` runs silently; everything is still checkpointed to
the run directory, so `report.md` and `code.R` tell the story afterwards.
